---
title: "Examine marker genes"
author: John Reid
output: html_document
---

`r date()`

```{r render, eval = FALSE, echo = FALSE}
#
# Execute this section to render this document.
#
devtools::load_all('..')
rmarkdown::render('marker-genes.Rmd')
```

```{r setup, echo = FALSE}
knitr::opts_chunk$set(fig.width=18, fig.height=12)
```

Load libraries.
```{r loadPkgs, message = FALSE}
library(ggplot2)
library(ggthemes)
petropoulos.colours <- scale_colour_manual(values = c("#00AE8A", "#CA8505", "#FF6C00", "#8586C3"))
```


# Marker genes

Define marker genes
```{r marker}
markers <- data.frame(
    gene = factor(
        c(
            'NANOG',   # EPI
            'NODAL',   # EPI
            'SOX2',    # EPI
            'DPPA5',   # EPI
            'GDF3',    # EPI
            'PRDM14',  # EPI
            'TDGF1',   # EPI
            'GATA4',   # PE
            'COL4A1',  # PE
            'FN1',     # PE
            'FGFR2',   # PE
            'LAMA4',   # PE
            'HNF1B',   # PE
            'PDGFRA',  # PE
            'SOX17',   # PE
            'GATA6',
            'GATA2',   # TE
            'GATA3',   # TE
            'DAB2'),   # TE
        levels = gene.levels),
    marker.lineage = factor(
        c(
            'EPI',
            'EPI',
            'EPI',
            'EPI',
            'EPI',
            'EPI',
            'EPI',
            'PE',
            'PE',
            'PE',
            'PE',
            'PE',
            'PE',
            'PE',
            'PE',
            NA,
            'TE',
            'TE',
            'TE')))
# devtools::use_data(markers, overwrite = TRUE)
# readr::write_csv(markers, '../extdata/Processed/markers.csv')
```

Plot distribution of expression values to determine thresholds of high
expression.
```{r plotMarker}
marker.expr <- expr.m %>% filter(gene %in% markers$gene)
ggplot(marker.expr, aes(x = log.rpkm.adj)) +
    geom_density() +
    geom_rug() +
    ylim(0, 3) +
    facet_wrap(~ gene) +
    theme_few()
```

Plot expression against inferred pseudotimes.
```{r exprVsPseudo}
ggplot(marker.expr %>% left_join(cell.meta),
       aes(x = pseudotime,
           y = log.rpkm.adj,
           colour = lineage)) +
    geom_point(alpha = .5) +
    facet_wrap(~ gene) +
    petropoulos.colours +
    theme_few()
ggsave(file.path('..', 'Plots', 'marker-pseudotime.pdf'))
```

Plot expression against capture times.
```{r exprVsCapture}
marker.capture <-
    marker.expr %>%
    left_join(cell.meta) %>%
    group_by(gene, lineage, capture) %>%
    summarise(
        obstime = obstime[1],
        mean.log.rpkm = mean(log.rpkm),
        sd.log.rpkm = sd(log.rpkm))
ggplot(marker.capture,
       aes(x = obstime,
           y = mean.log.rpkm,
           colour = lineage,
           ymin = mean.log.rpkm - sd.log.rpkm,
           ymax = mean.log.rpkm + sd.log.rpkm)) +
    geom_line() +
    geom_errorbar() +
    facet_wrap(~ gene) +
    petropoulos.colours +
    theme_few()
ggsave(file.path('..', 'Plots', 'marker-capture.pdf'))
```


# Most variable genes

Get the 500 most variable genes according to Sandberg & Lanner.
```{r mostVar}
most.var.genes <- filter(gene.variation, rank <= 500)$gene
stopifnot(500 == length(most.var.genes))
# devtools::use_data(most.var.genes)
```

Which marker genes are in the 500 most variable genes?
```{r markerMostVar}
markers %>% filter(gene %in% most.var.genes)
markers %>% filter(! gene %in% most.var.genes)
```


# Session information

`r date()` on `r Sys.info()[['nodename']]`.
```{r sessionInfo}
devtools::session_info()
```
