---
title: "Examine marker genes"
author: John Reid
output: html_document
---

`r date()`

```{r render, eval = FALSE, echo = FALSE}
#
# Execute this section to render this document.
#
devtools::load_all('..')
rmarkdown::render('marker-genes.Rmd')
```

Load libraries.
```{r loadPkgs, message = FALSE}
library(ggplot2)
library(ggthemes)
```


# Marker genes

Define marker genes
```{r marker}
markers <- data.frame(
    gene = factor(
        c(
            'NANOG',   # EPI
            'NODAL',   # EPI
            'SOX2',    # EPI
            'DPPA5',   # EPI
            'GDF3',    # EPI
            'PRDM14',  # EPI
            'TDGF1',   # EPI
            'GATA4',   # PE
            'COL4A1',  # PE
            'FN1',     # PE
            'FGFR2',   # PE
            'LAMA4',   # PE
            'HNF1B',   # PE
            'PDGFRA',  # PE
            'SOX17',   # PE
            'GATA6',
            'GATA2',   # TE
            'GATA3',   # TE
            'DAB2'),   # TE
        levels = gene.levels),
    lineage = factor(
        c(
            'EPI',
            'EPI',
            'EPI',
            'EPI',
            'EPI',
            'EPI',
            'EPI',
            'PE',
            'PE',
            'PE',
            'PE',
            'PE',
            'PE',
            'PE',
            'PE',
            NA,
            'TE',
            'TE',
            'TE')))
# devtools::use_data(markers, overwrite = TRUE)
```

Plot distribution of expression values to determine thresholds of high
expression.
```{r plotMarker}
marker.expr <- expr.m %>% filter(gene %in% markers$gene)
ggplot(marker.expr, aes(x = log.rpkm.adj)) +
    geom_density() +
    geom_rug() +
    ylim(0, 3) +
    facet_wrap(~ gene) +
    theme_few()
```

Plot expression against inferred pseudotimes.
```{r exprVsPseudo}
ggplot(marker.expr %>% left_join(cell.meta),
       aes(x = pseudotime, y = log.rpkm.adj, colour = lineage)) +
    geom_point(alpha = .5) +
    facet_wrap(~ gene) +
    scale_colour_few() +
    theme_few()
ggsave(file.path('..', 'Plots', 'marker-profiles.pdf'))
```


# Most variable genes


Get the 500 most variable genes according to Sandberg & Lanner.
```{r mostVar}
most.var.genes <- filter(gene.variation, rank <= 500)$gene
stopifnot(500 == length(most.var.genes))
# devtools::use_data(most.var.genes)
```

Which marker genes are in the 500 most variable genes?
```{r markerMostVar}
markers %>% filter(gene %in% most.var.genes)
markers %>% filter(! gene %in% most.var.genes)
```


# Session information

`r date()` on `r Sys.info()[['nodename']]`.
```{r sessionInfo}
devtools::session_info()
```
