---
title: "Estimate pseudotimes for Sandberg and Lanner's data"
author: John Reid
output: html_document
params:
    pseudotime.lineage: "epiblast"
    max.cells.per.capture: 500
    max.genes: 500
    num.held.out: 50
    inf.method: 'vb'
    sigma.tau: .25
    length.scale: 3
    seed: 1
---

`r date()`

```{r render, eval = FALSE, echo = FALSE}
#
# Execute this section to render this document.
#
# parameters = list(pseudotime.lineage = 'epiblast')
parameters = list(pseudotime.lineage = 'trophectoderm')
# parameters = list(pseudotime.lineage = 'primitive endoderm')
rmarkdown::render(
    'pseudotimes.Rmd',
    params = parameters,
    output_file = stringr::str_c('pseudotimes-', parameters$pseudotime.lineage, '.html'))
```

```{r setup, echo = FALSE}
pseudotime.lineage <- params$pseudotime.lineage
max.cells.per.capture <- as.integer(params$max.cells.per.capture)
max.genes <- as.integer(params$max.genes)
num.held.out <- as.integer(params$num.held.out)
inf.method <- params$inf.method
sigma.tau <- as.numeric(params$sigma.tau)
length.scale <- as.numeric(params$length.scale)
seed <- as.integer(params$seed)
set.seed(seed)
knitr::opts_chunk$set(fig.width=18, fig.height=12)
```

Load libraries.
```{r loadPkgs, message = FALSE}
library(ggplot2)
library(ggthemes)
library(dplyr)
library(reshape2)
library(stringr)
library(broom)
devtools::load_all()
devtools::load_all('/home/john/Dev/DeLorean')
```


# Filter cells in lineage

Show how many cells in each capture/lineage combination.
```{r numCells}
cell.meta %>% group_by(capture, lineage) %>% summarise(N = n())
```

Retrieve cells in lineage `r pseudotime.lineage` and subsample.
```{r}
cells.lineage <-
    cell.meta %>%
    filter('E3' != capture) %>%
    mutate(capture = factor(capture, levels = levels(capture), ordered = TRUE)) %>%
    filter(lineage %in% c('none', pseudotime.lineage)) %>%
    group_by(capture) %>%
    do(sample_n(., min(max.cells.per.capture, nrow(.)))) %>%
    ungroup()
```
Have `r nrow(cells.lineage)` cells.


# Find variable genes

Analyse which genes vary most across the capture time points and select just
the most variable.
```{r geneAov}
.aov <-
    cells.lineage %>%
    select(cell, embryonic.day) %>%
    left_join(expr.m) %>%
    group_by(gene) %>%
    do(tidy(aov(log.rpkm ~ embryonic.day, .))) %>%
    ungroup() %>%
    arrange(p.value)
genes.variable <- head(.aov, max.genes)$gene
genes.lineage <-
    gene.meta %>%
    filter(gene %in% genes.variable)
```

Build the expression matrix.
```{r exprMat}
expr <-
    expr.m %>%
    filter(gene %in% genes.lineage$gene) %>%
    filter(cell %in% cells.lineage$cell) %>%
    acast(gene ~ cell, value.var = 'log.rpkm')
expr <- expr[as.character(genes.lineage$gene), as.character(cells.lineage$cell)]
```

Build the `DeLorean` object.
```{r buildDL, eval=fit.model}
dl <- de.lorean(
    expr,
    genes.lineage,
    cells.lineage)
```
Re-level the cells by their capture time. This improves the ordering in later
plots.
```{r relevelCells}
dl$cell.meta$cell <- factor(
    dl$cell.meta$cell,
    levels=(cells.lineage %>% arrange(capture))$cell)
```


# Estimate hyperparameters

Examine data for empirical Bayes estimation of hyperparameters.
```{r empiricalBayes, eval=fit.model}
model.name <- 'lowrank'
dl <- estimate.hyper(
    dl,
    sigma.tau = sigma.tau,
    length.scale = length.scale,
    model.name = model.name)
```
```{r child=report.file("data")}
```
```{r child=report.file("hyper-parameters")}
```


# Fit model

```{r options}
num.inits <- default.num.cores()
```
Define and compile the model, find the best initialisation, and fit the model.
The model name is "`r model.name`", the inference method is "`r inf.method`"
with `r num.inits` initialisations, the seed is `r seed` and the data have
`r dim(dl)[1]` genes and `r dim(dl)[2]` cells.
```{r fitModel, eval=fit.model}
dl <- prepare.for.stan(dl)
dl <- compile.model(dl)
dl <- find.good.ordering(dl, seriation.find.orderings)
plot(dl, type='orderings')
dl <- pseudotimes.from.orderings(dl, num.to.keep=num.inits)
system.time(dl <- fit.model(dl, method=inf.method, num.inits=num.inits))
```
```
```{r examConv, eval=("sample" == inf.method)}
dl <- examine.convergence(dl)
```
```{r child=report.file("convergence"), eval=("sample" == inf.method)}
```


# Analyse posterior

Examine posterior and optimise best sample.
```{r posterior}
dl <- process.posterior(dl)
init.orderings.vs.pseudotimes.plot(dl) + scale_colour_few() + theme_few()
dl <- analyse.noise.levels(dl)
```
```{r child=report.file("posterior")}
```


# Profiles

Calculate expression profiles.
```{r makePredictions}
dl <- make.predictions(dl)
# plot(dl, type = 'profiles') + scale_colour_few() + theme_few()
```
```{r child=report.file("profiles")}
```


# Roughnesses

Get the held out genes.
```{r heldOut}
held.out.genes <- .aov$gene[(max.genes+1):(max.genes+num.held.out)]
held.out.expr <-
    expr.m %>%
    filter(gene %in% held.out.genes) %>%
    filter(cell %in% cells.lineage$cell) %>%
    acast(gene ~ cell, value.var = 'log.rpkm')
held.out.expr <- held.out.expr[as.character(held.out.genes), as.character(cells.lineage$cell)]
```

Calculate and plot the roughnesses of the held out genes.
```{r roughnesses}
dl <- roughness.test(dl, held.out.expr)
plot(dl, type = 'roughnesses')
plot(dl, type = 'profiles')
```


# Session information

`r date()` on `r Sys.info()[['nodename']]`.
```{r sessionInfo}
devtools::session_info()
```
